---
title: "BIS15L_AI_hmwk"
author: "Maya Ziebarth"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
This homework is a bit less traditional this week. There are some errors in the code below. In addition add notes and titles to the code chunks that explains what the code chunck does. Use AI tools, notes from class, or any other resources to help fix the errors. Feel free to edit any plots to your preference. Don't forget to answer the two questions at the bottom. Upon completion, you should not have to '#' out any of the code chunks.

Push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.   

Again make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  


```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```


1. load in the data 'blood_count_dataset.csv'
```{r}
blood_count <- read_csv("data/blood_count_dataset.csv") %>% clean_names()
```

2. Get an idea of what the data looks like
```{r}
glimpse(blood_count)
```

3. Find the maximum, minimum, and mean of the red bloods by gender
```{r}
blood_count %>% 
  group_by(gender) %>% 
  summarize(min=min(red_blood_cells), 
            mean=mean(red_blood_cells), 
            max=max(red_blood_cells)) 
```


4. Filter the data to only include males in the columns 3 through 9, and then calculate the column means for those columns.
```{r}
options(scipen = 999, digits = 2)

blood_count %>% 
  filter(gender == "Male") %>% 
  select(3:9) %>% 
  summarize_all(mean, na.rm = TRUE) 
```

5. Filter the data to only include males in the columns 3 through 9, and then calculate the column means for those columns.
```{r}
blood_count %>%
  filter(gender == "Female") %>%
  select(3:9) %>%
  summarize_all(mean, na.rm = TRUE)
```


6. Make a box plot of gender by the blood count tests
```{r}
blood_count %>%
  pivot_longer(cols = (3:9), names_to= "test", values_to = "values") %>%
  ggplot(aes(x = gender, y = values, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~test, scales = "free") 
  labs(title = "Boxplot of Blood Count Tests",
       x = "Test",
       y = "Values")
```

7. Group the ages into 20s, 30s, 40s, 50s, and 60s (60s and 70s should be in one group) and make a new column for those values. Make this a new dataframe called "grouped_bc".
```{r}
grouped_bc <- blood_count %>%
  mutate(age_group = case_when(
    age >= 20 & age <= 29 ~ "20s",
    age >= 30 & age <= 39 ~ "30s",
    age >= 40 & age <= 49 ~ "40s",
    age >= 50 & age <= 59 ~ "50s",
    age >= 60 & age <= 70 ~ "60s",
  ))

head(grouped_bc)
```

8. Make a boxplot from the new dataframe, 'grouped_bc', to show the blood count tests by age group.
```{r}
grouped_bc %>%
  pivot_longer(cols = (3:9), names_to= "test", values_to = "values") %>%
  ggplot(aes(x = age_group, y = values, fill = age_group)) +
  geom_boxplot() +
  facet_wrap(~test, scales = "free") +  # Facet by the "Test" variable
  labs(title = "Boxplot of Blood Count Tests by Age Group",
       x = "Age Group",
       y = "Values")
```

9. What are two drawbacks and two benefits of AI?
Drawbacks: 
1) It can lead you to rely on it and now on your skills, leaving you to not know how to funtion without it
2) It can be wrong and lead you to make mistakes
Benefits:
1) It can help you learn and understand things you may not have been able to before and help widen your skill set
2) Its great for troubleshooting 

10. Do you think you will use AI for coding in the future? Why or why not?
I'm conflicted about it. I want to use it but it is so far confusing me a bit more than helping me. I think once I get the hang of it, it will be a great tool to use only when needed, otherwise I will stick to my own coding. 


## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 