---
title: "Homework 7"
author: "Maya Ziebarth"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
``` 

## Data
**1. For this homework, we will use two different data sets. Please load `amniota` and `amphibio`.**  

`amniota` data:  
Myhrvold N, Baldridge E, Chan B, Sivam D, Freeman DL, Ernest SKM (2015). “An amniote life-history
database to perform comparative analyses with birds, mammals, and reptiles.” _Ecology_, *96*, 3109.
doi: 10.1890/15-0846.1 (URL: https://doi.org/10.1890/15-0846.1).
```{r}
amniota <- read.csv("data/amniota.csv") %>% clean_names()
```

`amphibio` data:  
Oliveira BF, São-Pedro VA, Santos-Barrera G, Penone C, Costa GC (2017). “AmphiBIO, a global database
for amphibian ecological traits.” _Scientific Data_, *4*, 170123. doi: 10.1038/sdata.2017.123 (URL:
https://doi.org/10.1038/sdata.2017.123).
```{r}
amphibio <- read.csv("data/amphibio.csv") %>% clean_names()
```

## Questions  
**2. Do some exploratory analysis of the `amniota` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  
```{r}
glimpse(amniota)
```